---
title: "Population genetics"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    self-contained: true
editor_options: 
  chunk_output_type: console
execute: 
  warning: false
  echo: true
---

# Project Setup

```{r}
#| label:  setup

cran <- c("devtools", "poppr", "vcfR", "ape", "RColorBrewer")
git <- c("hemstrow/snpR") 

suppressPackageStartupMessages({
  for (pkg in cran) { 
    if( !require(pkg, character.only = TRUE) ) {
      install.packages(pkg)
      library(pkg, character.only = TRUE)
    } 
  }
  
  for (pkg in git) {
    if( !require(sub(".*/", "", pkg), character.only = TRUE) ) {
      devtools::install_github(pkg, upgrade = TRUE)
      library(sub(".*/", "", pkg), character.only = TRUE)
    }
  }
}); remove(cran, git, pkg)

source("https://inkaverse.com/docs.r")
```

# References

- https://grunwaldlab.github.io/poppr/index.html
- https://grunwaldlab.github.io/Population_Genetics_in_R/index.html
- https://grunwaldlab.github.io/Population_Genetics_in_R/gbs_analysis.html
- https://github.com/hemstrow/snpR
- https://www.nature.com/articles/s41437-022-00535-z
- https://github.com/hemstrow/snpR/blob/master/vignettes/snpR_introduction.Rmd

# Poppr 

Poppr is an R package designed for analysis of populations with mixed modes of sexual and clonal reproduction. 

> There are several stand-alone programs that can handle these types of data sets, but they are often platform specific and often only accept specific data types. Furthermore, a typical analysis often involves switching between many programs, and converting data to each specific format.

Poppr allows analysis of haploid and diploid dominant/co-dominant marker data including microsatellites, Single Nucleotide Polymorphisms (SNP), and Amplified Fragment Length Polymorphisms (AFLP). 

# Caso de estudio

Data set of 262 Prunus samples based on the study by Cho et al. (2021) on Prunus takesimensis and related taxa. The samples include individuals from P. takesimensis, P. sargentii, and other related groups, and were analyzed from a VCF file containing 16,137 SNP variants. This dataset was used in R to assess genetic diversity, population structure, genetic distances, and clustering patterns among populations.

## Import data

```{r}
prunusVCF <- read.vcfR("Data_5_262_Prunus.vcf")
prunusVCF

library(readxl)

popdata <- read_excel("popdata_prunus_262.xlsx")

all(colnames(prunusVCF@gt)[-1] == popdata$AccessID)

glprunus <- vcfR2genlight(prunusVCF) # data set into an object that is usable by poppr


```

The vcfR2genlight function subsets the data to filter loci that are not bi-allelic, returning an object that contains only loci with two alleles.

We are required to specify the ploidy of the organism in order to calculate some population genetic metrics. P. rubi is a diploid organism, so we will specify a ploidy of two.

```{r}
ploidy(glprunus) <- 2
```

Our biological question requires predetermined populations. We can add them to the genlight object as part of the pop (population) slot. In order to specify the population, we added the State column from our pop.data data frame to the pop slot of our genlight object:

```{r}
pop(glprunus) <- popdata$State
glprunus
```

# Population genetic analyses

When samples are very similar, they are grouped by short branches. The longer the branch, the higher the number of substitutions and the higher the genetic distance is between samples or clusters.

For this tutorial, we will build a distance tree to obtain an initial assessment of the population structure of the P. rubi samples in the western US. We will reconstruct a distance tree based on the UPGMA algorithm, with 100 bootstrap replicates to assess branch support:

```{r}
tree <- aboot(glprunus
               , tree = "upgma"
               , distance = bitwise.dist
               , sample = 100
               , showtree = F, cutoff = 50, quiet = T)
```

## Distance tree

```{r}
cols <- brewer.pal(n = nPop(glprunus), name = "Dark2")
plot.phylo(tree, cex = 0.25, font = 1, adj = 0, tip.color = cols[pop(glprunus)])
bs <- as.numeric(tree$node.label) 
nodelabels(ifelse(!is.na(bs) & bs >= 90, bs, ""),
           adj = c(1.2, -0.2),
           frame = "n",
           cex = 0.5,
           font = 2,
           xpd = TRUE)
#legend(35,10,c("CA","OR","WA"),cols, border = FALSE, bty = "n")
legend("topleft", legend = levels(pop(glprunus)), fill = cols, border = FALSE, bty = "n", cex = 0.8)
axis(side = 1)
title(xlab = "Genetic distance (proportion of loci that are different)")
```

## PCA
# le puse solo 3000 porque sino se demoraba mucho mi computadora
```{r}
prunus.pca <- glPca(glprunus[, 1:3000], nf = 3)

prunus.pca.scores <- as.data.frame(prunus.pca$scores)
prunus.pca.scores$pop <- pop(glprunus)

plot <- ggplot(prunus.pca.scores, aes(x=PC1, y=PC2, colour=pop)) + 
  geom_point(size=2) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  scale_color_manual(values = cols)+ 
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_bw()
plot
```

## DAPC

The DAPC is a multivariate statistical approach that uses populations defined a priori to maximize the variance among populations in the sample by partitioning it into between-population and within-population components. DAPC thus maximizes the discrimination between groups. 

# acá también solo puse 100 porque si no se lageaba mi pc


```{r}
prunus.dapc <- dapc(glprunus[, 1:1000], n.pca = 3, n.da = 2)

compoplot(prunus.dapc, col = cols, posi = "top")
```

# snpR

```{r}
glprunus_snpR <- snpR::convert_vcfR("Data_5_262_Prunus.vcf")
glprunus_snpR

popdata <- read_excel("popdata_prunus_262.xlsx")

sample.meta(glprunus_snpR)$pop <- popdata$State

sample.meta(glprunus_snpR)
snp.meta(glprunus_snpR)
```

## Statistics

```{r}
glprunus_snpR <- calc_basic_snp_stats(x = glprunus_snpR
                                      , facets = "pop"
                                      , sigma = 200
                                      , step = 50
                                      , par = 8
                                      )

stat <- get.snpR.stats(glprunus_snpR, "pop")
stat %>% kable()

snpR::plot_pairwise_fst_heatmap(glprunus_snpR, facets = "pop")

```

## PCA

```{r}
plot <- plot_clusters(glprunus_snpR, facets = c("pop"))
plot
```

## Genetic distance

```{r}
gd <- snpR::calc_genetic_distances(glprunus_snpR, facets = c("sampID"), method = "Edwards")
dat <- get.snpR.stats(gd, facets = "sampID", stats = "genetic_distance")
heatmap(as.matrix(dat$sampID$.base$Edwards))
```

## Population structure

```{r}
snpR::plot_structure(glprunus_snpR, facet = c("pop"), k = 2:4,
                     clumpp = FALSE, method = "snapclust")

snpR::plot_structure(glprunus_snpR, facet = c("pop"), k = c(4))
```

